---
title: "Genetic models: writing, generating and measuring QTL effects"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Genetic models: writing, generating and measuring QTL effects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

A trait's genetics in `tidybreed` is a set of **genome-effect terms**: additive,
dominance and additive-by-additive (A x A) coefficients on QTL, stored in the
`genome_effects` tables and evaluated into `ind_tgv` by `add_tgv()`. There are
four ways in, one per job:

| Job | Function |
|---|---|
| You know the coefficients | `define_genome_effect_terms()` with `ad_terms()`, `aa_terms()`, `genotype_terms()` |
| Sample additive effects that hit a target `G` exactly | `define_additive_effects()` |
| Sample additive + dominance + A x A effects that hit `G_A`, `G_D`, `G_AA` exactly | `define_genome_effects()` |
| Measure what a population actually has | `extract_genetic_variance()` |

The two generators sample **and** calibrate: their effects are written under the
reserved owner `"generated"`, which means "scaled to the stored target". Known
coefficients always go through the writer, under your own owner.

## A small population

Everything below runs on a 60-locus genome held in memory.

```{r pop, message = FALSE}
library(tidybreed)
library(dplyr)

set.seed(1)
pop <- open_pop(pop_name = "models", db_name = ":memory:") |>
  define_genome(n_loci = 60, n_chr = 2, chr_len_Mb = 100) |>
  define_founder_haplotypes(n_haplotypes = 400)
pop <- pop |> get_table("founder_haplotypes") |>
  add_founders(n_males = 100, n_females = 100, line_name = "A")
for (t in c("ADG", "BF", "WW", "FI", "LS")) pop <- define_trait(pop, t)
```

## 1. Known coefficients: the writer

With functional coding, `a` is half the homozygote difference, `d` the
heterozygote's deviation and `e` the A x A coefficient on the centred dosages.
`p` is required: it is the basis of the reported mean. The builders only build
rows; `define_genome_effect_terms()` writes them.

```{r writer}
pop <- pop |>
  define_genome_effect_terms(
    trait_name = "ADG",
    terms = rbind(
      ad_terms(locus_name = c("Locus_10", "Locus_44"),
               a = c(0.30, -0.12), d = c(0.10, 0.05), p = c(0.35, 0.60),
               coding = "functional"),
      aa_terms(locus_name_1 = "Locus_10", locus_name_2 = "Locus_44",
               e = 0.08, p_1 = 0.35, p_2 = 0.60, coding = "functional")))

pop <- pop |> get_table("ind_meta") |> add_tgv("ADG")
get_table(pop, "ind_tgv") |> collect() |> arrange(id_ind) |> head(6)
```

Each individual gets one value per component; the `ind_tgv_total` view adds
them. Functional coding stores the heterozygote deviation as an `indicator`
term on the heterozygous genotype, so the components are the functional ones
(`additive`, `indicator`, `interaction`), not breeding values: the breeding
value of this model also carries part of `d` and `e`. Nothing here is
calibrated to a variance: a hand-written model has the variance its
coefficients give.

## 2. Additive effects to a target: `define_additive_effects()`

Pass a target `G` and the sampled effects are rescaled so the additive
covariance is `G` **exactly** under the anchor: `"genic"` (the HWE expectation
at the base allele frequencies, default) or `"realised"` (the base individuals'
own genotypes). The target is stored in `trait_var_comp` with the effects, in
one transaction.

```{r additive}
G <- matrix(c(1.0, 0.4,
              0.4, 2.0), 2, dimnames = list(c("BF", "WW"), c("BF", "WW")))
set.seed(2)
pop <- get_table(pop, "genome_meta") |>
  filter(locus_id <= 40) |>
  define_additive_effects(c("BF", "WW"), G = G)
```

The message reports what other populations would see (here the founder pool,
with the sampling LD of 400 haplotypes). `line_name =` writes a line-specific
variant calibrated to that line's own target, for crossbreeding designs.

## 3. Additive, dominance and epistasis: `define_genome_effects()`

The same idea with three targets. Dominance degrees are drawn around
`dominance_degree_mean` (0.19 by default), A x A effects sit on random pairs of
QTL (or the `pairs` you supply), and all three blocks are calibrated together,
highest order first.

```{r genome}
set.seed(3)
pop <- get_table(pop, "genome_meta") |>
  define_genome_effects("FI", G_A = 1, G_D = 0.3, G_AA = 0.2)
```

The messages report the additive floor of this draw (below), the inbreeding
depression the dominance effects imply (`inbreeding_depression =` sets it
instead), and the founder pool's view. A re-run replaces the trait's
`"generated"` terms and never touches terms you wrote yourself.

**The additive floor.** Dominance and epistasis induce additive effects of
their own (an allele substitution changes the dominance and pair values too),
so the additive variance cannot be made arbitrarily small. A `G_A` below what
the drawn architecture can cancel is refused, with the floor named:

```{r floor, error = TRUE}
set.seed(4)
pop <- get_table(pop, "genome_meta") |>
  filter(locus_id <= 10) |>
  define_genome_effects("LS", G_A = 0.001, G_D = 1)
```

The floor belongs to the sampled architecture: another seed, or more loci,
gives another.

## 4. Measuring: `extract_genetic_variance()`

The extractor decomposes the stored model on any group of individuals, under
the same two anchors, and writes nothing. Its rows use `trait_var_comp`'s
column names, so a target comparison is a join.

```{r measure}
measured <- get_table(pop, "ind_meta") |>
  extract_genetic_variance(c("BF", "WW"), anchor = "realised")
measured

targets <- get_table(pop, "trait_var_comp") |> collect()
inner_join(measured, targets,
           by = c("effect_name", "trait_name_1", "trait_name_2"),
           suffix = c("_measured", "_target")) |>
  select(effect_name, trait_name_1, trait_name_2,
         cov_value_measured, cov_value_target)
```

The generation targets were calibrated under the genic anchor at the founder
pool's frequencies; these 200 individuals are a finite sample of it, so the
realised values differ a little. `anti_join(targets, measured, ...)` lists
targets with no measured block (a trait without terms, say).

## What "exact" promises

- A generator's result is exact for a feasible target **under the named anchor
  and the selected QTL**: other populations, later generations and other
  anchors see what their frequencies and LD give. Every call reports one such
  comparison.
- A line-scoped call calibrates its own variant only.
- `method = "union"` (each trait its own QTL set) delivers the variances
  exactly but the covariances only approximately, with a warning that gives
  them; `method = "shared"` is exact.
- One call writes one `parent_origin` scope.
- The extractor measures the population **you select**. Compare like with
  like: the generation anchor, base and scope.

```{r close, include = FALSE}
close_pop(pop)
```
