---
title: "Introduction to tidybreed"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Introduction to tidybreed}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

## What `tidybreed` is

`tidybreed` simulates animal and plant breeding programs. You build a population
step by step — a genome, founder animals, traits, phenotypes, matings, selection —
and the package tracks every individual, haplotype, and record for you.

What makes it different from other simulators is where the data lives. **Everything
is stored in a [DuckDB](https://duckdb.org/) database, not in R memory.** A
simulation with a million animals does not need a million animals' worth of RAM.
The database is a plain file on disk, so runs are resumable, shareable, and
queryable with ordinary SQL long after the R session that made them has ended.

The API is pipe-friendly and built on `dplyr` verbs, so if you know the tidyverse
you already know most of the syntax.

This vignette walks through a complete single-generation simulation. Every code
block below actually runs when this page is built, so what you see is real output.

## Installation

`tidybreed` is not on CRAN yet. Install from GitHub — `pak` handles the compiled
code and dependencies cleanly:

```{r install, eval = FALSE}
install.packages("pak")
pak::pak("austin-putz/tidybreed")
```

You will need a C++ compiler (Xcode command line tools on macOS, Rtools on
Windows). Then:

```{r libs, message = FALSE}
library(tidybreed)
library(dplyr)

set.seed(1)   # so this page renders the same numbers every time
```

## How the package works

Three ideas explain almost all of the API. They are worth reading before the code.

### 1. R is the control layer, DuckDB is the storage layer

Your R session holds a small object with a database connection in it. It does
**not** hold your genotypes. When you ask for a table you get a *lazy reference* —
nothing is pulled into R until you call `collect()`. This is why you should filter
first and collect last.

### 2. There are two object types

| Object | Created by | What it is |
|---|---|---|
| `tidybreed_pop` | `open_pop()` | The population: a database connection plus a table registry |
| `tidybreed_table` | `get_table()` | A lazy reference to one table, plus any pending filter |

Most functions take a `tidybreed_pop` and return a `tidybreed_pop`, so they chain:

```r
pop <- pop |> define_genome(...) |> define_founder_haplotypes(...)
```

But **action functions** — the ones that operate on a *subset of rows* — take a
`tidybreed_table` as their first argument instead. That is how you tell them which
rows to act on:

```r
pop <- pop |>
  get_table("ind_meta") |>       # which table
  filter(sex == "F") |>          # which rows
  add_phenotype("ADG")           # what to do
```

`add_founders()`, `add_phenotype()`, `add_tbv()`, `add_index()`,
`define_additive_effects()`, `define_chip()`, `add_genotypes()`,
`extract_genotypes()`, and `mutate_table()` all work this way. Most of them return
a `tidybreed_pop`, so the chain continues; the `extract_*` family is the exception
— it hands you data back as a tibble, since pulling results into R is the whole
point of it.

There is no positional row selection anywhere in `tidybreed` — no `TRUE`/`FALSE`
masks, no integer indices. Database row order is not guaranteed, so selection
always goes through names (`id_ind`, `locus_name`) or a SQL filter.

### 3. Genetics and observations are separate layers

This one trips people up, so it is worth being explicit. A trait's **genetics** and
the **phenotype someone writes on a clipboard** are defined by two different
functions:

| You want to specify | Function | Examples |
|---|---|---|
| Genetic architecture | `define_trait()` | `target_add_var`, `target_add_mean`, `expressed_parent`, `units` |
| What gets observed | `define_phenotype()` | `mean`, `type`, `residual_var`, `expressed_sex`, `repeatable` |

For a simple trait the two share a name and you call both. The payoff comes later:
one observed phenotype (weaning weight) can be built from several genetic
components (direct + maternal), and the split makes that natural instead of a
special case.

## Open a population

`open_pop()` creates the database. Here we use `":memory:"` so this vignette writes
nothing to disk; in a real run you would omit `db_name` and get a `sim.duckdb` file
inside an organised output folder.

```{r open}
pop <- open_pop(pop_name = "demo", db_name = ":memory:")

pop
```

Printing a population gives you a live summary. Right now there is almost nothing
in it — sections appear as you add data.

## Define the genome

```{r genome}
pop <- pop |>
  define_genome(
    n_loci     = 500,   # total loci (SNP and QTL both come from this pool)
    n_chr      = 5,     # chromosomes
    chr_len_Mb = 100    # length of each chromosome in megabases
  )

pop |> get_table("genome_meta")
```

`genome_meta` holds one row per locus with its **physical** position (`pos_bp`).
The **genetic** map lives in its own table, because a genome can have several maps
— male vs female, or line-specific:

```{r genome_map}
pop |> get_table("genome_map")
```

Per-chromosome inheritance rules live in `chr_meta`. Every chromosome starts as a
normal diploid autosome; `define_chr()` is how you would declare an X, Y, or
mitochondrial chromosome.

```{r chr_meta}
pop |> get_table("chr_meta") |> collect()
```

## Explore the database

Two helpers exist so you never have to guess what is in the database.
`schema(pop)` lists every table:

```{r schema}
schema(pop)
```

`describe_table()` explains one table's columns:

```{r describe}
describe_table(pop, "ind_meta")
```

Remember that `get_table()` is lazy. The filter below runs *inside* DuckDB and only
the answer comes back to R:

```{r lazy}
pop |>
  get_table("genome_meta") |>
  filter(chr == 1L) |>
  count() |>
  collect()
```

## Create founder haplotypes

Founders are sampled from a pool of haplotypes. Tagging the pool with a
`line_name` is good practice — it is what lets you run crossbreeding later.

```{r haps}
pop <- pop |>
  define_founder_haplotypes(
    line_name       = "A",
    n_haplotypes    = 100,
    method          = "uniform",
    min_allele_freq = 0.05,
    max_allele_freq = 0.95
  )
```

Several methods control the allele-frequency spectrum and linkage disequilibrium:

| `method` | What it gives you |
|---|---|
| `"uniform"` | Frequencies drawn uniformly between `min_allele_freq` and `max_allele_freq` |
| `"fixed"` | Every locus at the same `allele_freq` |
| `"beta"` | Beta-distributed frequencies (`beta_shape1`, `beta_shape2`) — U-shaped is realistic |
| `"balding_nichols"` | FST-based divergence between lines (`fst`, `mean_freq`) |
| `"mosaic"` | Haplotypes built from templates, producing simple LD (`n_templates`, `switch_rate`) |
| `"gaussian_copula"` | Correlated frequencies with distance-based decay (`decay_rate`) |

## Add founder individuals

This is the first action function, and it shows why the `tidybreed_table` pattern
exists: you pipe in the haplotype pool you want to sample *from*.

Any extra named argument — like `gen` below — becomes a new column on `ind_meta`,
written in the same transaction as the animals themselves.

```{r founders}
pop <- pop |>
  get_table("founder_haplotypes") |>
  filter(line_name == "A") |>
  add_founders(
    n_males   = 250,
    n_females = 250,
    line_name = "A",
    gen       = 0L      # custom column
  )

pop |> get_table("ind_meta") |> collect() |> head()
```

> **Watch the type suffix.** `gen = 0L` creates an `INTEGER` column; `gen = 0`
> would create a `DOUBLE`. The same applies to `NA_integer_`, `NA_character_`,
> and friends.

Haplotypes are stored in long format — one row per individual × parent × locus:

```{r haplo}
pop |> get_table("ind_haplotype")
```

## Add your own columns

`mutate_table()` adds or updates columns on *any* table. This is the mechanism for
keeping simulation state in the database instead of in parallel R objects.

```{r mutate}
# Broadcast a value to every row
pop <- pop |> get_table("ind_meta") |> mutate_table(farm = "Iowa")

# Update only some rows
pop <- pop |>
  get_table("ind_meta") |>
  filter(sex == "M") |>
  mutate_table(farm = "AI_Stud")

pop |> get_table("ind_meta") |> count(sex, farm) |> collect()
```

The warning above is deliberate: `mutate_table()` tells you whenever it overwrites
values that were already there, so you notice an accidental clobber instead of
discovering it three generations later.

Calling `mutate_table()` on an *empty* table still creates the column, which lets
you declare a typed schema before any data arrives:

```r
pop |> get_table("ind_ebv") |> mutate_table(model_version = NA_character_)
```

## Define a trait

The genetic layer. `target_add_var` is the additive genetic variance the QTL
effects will be scaled to hit.

```{r trait}
pop <- pop |>
  define_trait("ADG", target_add_var = 0.25, units = "kg/day")

pop |> get_table("trait_meta") |> collect()
```

Note what is *not* here: no mean, no residual variance, no trait type. Those are
observation-layer properties and come later.

## Assign QTL effects

Which loci are QTL is decided by the filter you pipe in. Here chromosomes 4 and 5
carry the QTL, leaving 1–3 free for a SNP chip.

```{r qtl}
pop <- pop |>
  get_table("genome_meta") |>
  filter(chr %in% c(4L, 5L)) |>
  define_additive_effects("ADG")

pop |> get_table("genome_effects") |> collect() |> head()
```

QTL membership is **implicit**: a locus is a QTL for a trait if it has a row in
`genome_effects`. There is no `is_QTL` flag to keep in sync. `base_allele_freq` is
stored alongside each effect because it is what centres breeding values.

## Define the phenotype

The observation layer. With `target_add_var = 0.25` and `residual_var = 0.75`,
this trait has a heritability of 0.25.

```{r pheno_def}
pop <- pop |>
  define_phenotype(
    "ADG",
    type         = "continuous",
    mean         = 1.0,
    residual_var = 0.75
  )

pop |> get_table("phenotype_meta") |> collect() |> select(1:6)
```

`type` also accepts `"count"`, `"categorical"` (with `prevalence` or `thresholds`),
and `"derived_formula"`.

## Add a fixed effect

Non-genetic terms attach to the *phenotype*, not the trait. Here males grow
0.30 kg/day faster than females, with females as the reference level.

```{r fixed}
pop <- pop |>
  define_effect_fixed_class(
    "ADG",
    effect_name   = "sex",
    source_column = "sex",
    levels        = c(M = 0.30, F = 0)
  )
```

`define_effect_fixed_cov()` adds a regression on a covariate, and
`define_effect_random()` adds a named random effect such as pen or litter.

## Compute breeding values and phenotypes

`add_tbv()` computes true breeding values — the simulation's ground truth, which
you would never know in a real population.

```{r tbv}
pop <- pop |> get_table("ind_meta") |> add_tbv("ADG")

pop |> get_table("ind_tbv") |> collect() |> head()
```

`add_phenotype()` builds the observed record: intercept, plus fixed and random
effects, plus the breeding value, plus a sampled residual. It calls `add_tbv()`
internally, so you can go straight to it.

```{r pheno}
pop <- pop |> get_table("ind_meta") |> add_phenotype("ADG")

pop |> get_table("ind_phenotype") |> collect() |> head()
```

The sex effect we defined should show up as a gap of roughly 0.30 between the
group means:

```{r pheno_summary}
pop |>
  get_table("ind_phenotype") |>
  collect() |>
  left_join(
    pop |> get_table("ind_meta") |> collect() |> select(id_ind, sex),
    by = "id_ind"
  ) |>
  group_by(sex) |>
  summarise(n = n(), mean_ADG = mean(pheno_value))
```

## A second, genetically correlated trait

Real programs select on several traits at once. Define the second trait, then
supply a genetic covariance matrix.

```{r trait2}
pop <- pop |> define_trait("BF", target_add_var = 0.30, units = "mm")

G <- matrix(
  c(0.25, 0.08,
    0.08, 0.30),
  nrow = 2,
  dimnames = list(c("ADG", "BF"), c("ADG", "BF"))
)

pop <- pop |> define_effect_cov_matrix("gen_add", G)

pop |> get_table("trait_var_comp") |> collect()
```

Covariance matrices are stored in the database, not passed around in R. Now one
call draws correlated effects for both traits from `MVN(0, G)`:

```{r qtl2}
pop <- pop |>
  get_table("genome_meta") |>
  filter(chr %in% c(4L, 5L)) |>
  define_additive_effects(c("ADG", "BF"))

pop <- pop |>
  define_phenotype("BF", type = "continuous", mean = 12, residual_var = 0.70)
```

> This **replaces** the ADG effects assigned earlier with a fresh correlated draw.
> Re-running `define_additive_effects()` for a trait always overwrites its rows in
> `genome_effects`, so recompute anything derived from them.

```{r tbv2}
pop <- pop |> get_table("ind_meta") |> add_tbv()   # no trait_name = all traits

pop |>
  get_table("ind_tbv") |>
  collect() |>
  group_by(trait_name) |>
  summarise(n = n(), mean_tbv = mean(tbv_value), var_tbv = var(tbv_value))
```

The realised variances are in the neighbourhood of the 0.25 and 0.30 we asked
for, but they are not exact — and that is worth understanding rather than
glossing over. `target_add_var` scales the QTL effects so the additive variance
comes out right **in the founder haplotype pool**. The animals you actually
created are a finite sample from that pool, so their realised variance wobbles
around the target. Larger founder populations wobble less.

## Make some offspring

Matings are described by an ordinary tibble: one row per offspring, with
`id_parent_1`, `id_parent_2`, `sex`, and `line_name` required. Any custom
`ind_meta` column can ride along.

```{r matings}
sires <- pop |> get_table("ind_meta") |> filter(sex == "M") |> collect() |> pull(id_ind)
dams  <- pop |> get_table("ind_meta") |> filter(sex == "F") |> collect() |> pull(id_ind)

matings <- tibble::tibble(
  id_parent_1 = rep(sires[1:5], each = 4),
  id_parent_2 = dams[1:20],
  sex         = rep(c("M", "F"), 10),
  line_name   = "A",
  gen         = 1L
)

head(matings)
```

`add_offspring()` simulates meiosis — crossovers are drawn from the genetic map in
`genome_map`, respecting the per-chromosome rules in `chr_meta`.

```{r offspring}
pop <- add_offspring(pop, matings)

pop |> get_table("ind_meta") |> filter(gen == 1L) |> collect() |> head()
```

Because you build the mating plan yourself, any design is possible — nested,
factorial, reciprocal crosses, assortative mating — without the package needing a
special argument for each.

```{r pop_now}
pop
```

## SNP chips and genotypes

A chip is a named set of loci. Mark them with a filter:

```{r chip}
pop <- pop |>
  get_table("genome_meta") |>
  filter(chr %in% c(1L, 2L, 3L)) |>
  define_chip("50K")

pop |> get_table("genome_meta") |> count(is_50K) |> collect()
```

`add_genotypes()` records *which animals were genotyped* — here, only the
offspring:

```{r genotype}
pop <- pop |>
  get_table("ind_meta") |>
  filter(gen == 1L) |>
  add_genotypes("50K")

pop |> get_table("ind_meta") |> count(has_50K) |> collect()
```

`extract_genotypes()` returns the familiar wide 0/1/2 dosage matrix, ready for
GBLUP or export:

```{r extract}
geno <- pop |> get_table("ind_meta") |> extract_genotypes("50K")

dim(geno)
geno[1:5, 1:6]
```

Haplotypes are the source of truth and dosages are derived on demand, so nothing
is stored twice. If you need dosages repeatedly, `add_dosage()` caches them in
`ind_genotype`.

## Selection index

Register the weights:

```{r index_def}
pop <- pop |>
  define_index("terminal",
               trait_names = c("ADG", "BF"),
               index_wts   = c(1.2, -0.8))

pop |> get_table("index_meta") |> collect()
```

The rows with a missing `index_name` are the global economic-weight entries that
`define_trait()` writes for every trait; the `terminal` rows are our index.

Because we know the true breeding values, we can compute the **true** index —
useful for measuring how well selection actually worked:

```{r true_index}
pop <- pop |> get_table("ind_meta") |> add_tbv(index_names = "terminal")

pop |> get_table("ind_true_index") |> collect() |> head()
```

`add_index()` does the same arithmetic on any table of values. In a real program
you would run it on `ind_ebv` after a BLUP evaluation; here we use the TBVs.
**Filter first** so there is exactly one value per individual per trait:

```{r add_index}
pop <- pop |>
  get_table("ind_tbv") |>
  filter(trait_name %in% c("ADG", "BF")) |>
  add_index("terminal")

pop |>
  get_table("ind_index") |>
  collect() |>
  arrange(desc(index_value)) |>
  head(5)
```

Those top animals are your selection candidates. Pull their IDs, build a new
mating plan, call `add_offspring()`, and you have closed the loop on a breeding
cycle.

## Close the population

```{r close}
close_pop(pop)
```

For an on-disk run, `close_pop()` flushes everything to the `.duckdb` file and you
can pick the simulation back up later:

```r
pop <- restore_pop("tidybreed_output/sim.duckdb")
```

Nothing needs to be re-specified — the genome, traits, variance components, and
every animal are already in the file. This is the payoff of keeping configuration
in the database rather than in R.

## Where to go next

This vignette covered the main path. The package has more, grouped by prefix:

| Prefix | Meaning |
|---|---|
| `open_` / `restore_` | Create or reopen a population |
| `define_` | Write model configuration — how the simulation should behave |
| `add_` | Write simulation output — data the model produced |
| `mutate_` | Add or update columns on an existing table |
| `get_` / `extract_` | Read data out, lazily or in analysis format |
| `remove_` / `archive_` | Delete rows, or stamp a finished replicate |

Functions not shown above, with a pointer to their help page:

| Function | What it does |
|---|---|
| `?add_ebv` | Run BLUPF90 (or parent average) and store estimated breeding values |
| `?define_chr` | Sex chromosomes, organelles, and achiasmatic meiosis |
| `?define_phenotype` | Composite traits: maternal effects and social genetic effects, via `components` |
| `?define_residual_cov` | Correlated or heterogeneous residual (co)variances |
| `?define_effect_random` | Named random effects such as pen, litter, or herd-year-season |
| `?define_effect_fixed_cov` | Fixed regression on a covariate |
| `?define_trait_simple` | Shortcut wrapper: `define_trait()` + `define_additive_effects()` |
| `?archive_replicate` | Collect many replicates into one archive database |
| `?remove_rows` | Delete rows safely across related tables |
| `?mutate_group_seq` | Litter and group utilities (`mutate_group_*`) |
| `?define_table` | Create your own tables inside the population |

The project README covers installation, global options, the output directory
layout, and scenario YAML files in more detail.
