---
title: "Introduction to tidybreed"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to tidybreed}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

`tidybreed` simulates animal and plant breeding programs using a pipe-friendly
API backed by a file-based DuckDB database.

## Installation

Eventually, I will try to push the package to CRAN, however, right now the
package is only available via GitHub, see below. 

```{r install_cran, eval=FALSE}
# install.packages("tidybreed")
```

I highly suggest using the `pak` package to download `tidybreed` from GitHub. 

```{r install_github, eval=FALSE}
# install pak package
install.packages("pak", repos = "https://packagemanager.posit.co/cran/latest")

# loadd pak
library(pak)

# install this package
pak::pak("austin-putz/tidybreed")
``` 


## Basic usage

First, make sure to load `tidybreed`. 

```{r load_tidybreed, eval=FALSE}
library(tidybreed)
```

Users should refer to the `/vignettes/qmd/` folder and open the HTML file
to learn this package right now. Later I will build this official `.Rmd` file
for users once I hit version `1.0.0` and get something official. 



